Quality Criteria for SKOS Vocabularies

### 2012-03-22 criteria.tex
finished labeling and documentation issues

\section{Quality Criteria for SKOS Vocabularies}\label{sec:criteria}

@@

We identified an initial set of quality criteria for SKOS vocabularies by reviewing existing literature and manually examining existing vocabularies. We published our findings online\footnote{\url{https://github.com/cmader/qSKOS/wiki/Quality-Criteria-for-SKOS-Vocabularies}} and requested feedback from experts via public mailing list and informal face to face discussions. Based on the received responses, we then selected a subset of these criteria and formalized them into computable \textbf{quality checking functions}. Each function operates on a SKOS vocabulary, which we define, for the purpose of this work, as follows:

\begin{definition}[SKOS Vocabulary] Let a SKOS vocabulary be a tuple of the form $V = \langle IR, C, SR, ULV \rangle$, with \(IR = I_{CEXT}(\texttt{rdfs:Resource}^\mathcal{I})\) being the set of \textbf{resources}, \(C = I_{CEXT}(\texttt{skos:Concept}^\mathcal{I})\) being the set of \textbf{conceptual resources}, \(SR = I_{EXT}(\texttt{skos:semanticRelation}^\mathcal{I})\) being the set of \textbf{semantic relations} associating conceptual resources with one another, and $ULV \subseteq I_{CEXT}(\texttt{rdfs:Literal}^\mathcal{I})$ being the set of \textbf{untyped literal values}. Further, we let $V$ be the fully entailed RDFS interpretation~\cite{RDFSEM2012} of the underlying RDF graph, enriched by entailment of \texttt{owl:inverseOf} properties.

@@

% @CM Are you sure about the term "untyped literal"? The RDF spec talkes about
% plain literals
% from the RDF simple interpretation: "6. A distinguished subset LV of IR, called the set of literal values, which contains all the plain literals in V"; "all plain literals" which doesn't state "only plain literals", so I think we need the subset here; Also, Hitzler defines LV as the set of literal values, "containing (at least) all untyped literals in V", which indicates that LV also contains typed literals.

The commonality of all quality checking functions is that they can be applied on a given SKOS vocabulary and report a (nested) set of vocabulary resources, being the cause of anticipated quality degradations. Therefore, we can formalize a quality criterion as being an abstract quality checking function $f:V \rightarrow \mathcal{P}(IR)$, mapping a given vocabulary $v \in V$ to the powerset of its resources. Each quality criterion we found can be implemented as a realization of such an abstract filtering function, possibly with additional parameters. Statistical analysis can be computed on top of the return values.

\todo{CM}{the term criterion still sounds strange. We must somehow clarify what passing/failing a criterion means there.}
\todo{BH}{maybe we should use the term "criteria" for our initial quality findings published online and only speak of "quality functions" in this publication. A criterion implies a strict measure (pass/fail) whereas our functions only provide "hints" for later human judgement. they are "computable guidelines"}

% Structure for each criterion description
%- origin (refs) and design rationale
%- formal definition that specifies how to compute the metric

In the following, we explain the origins and design rationale for each quality criterion (QC) and describe how we implemented the corresponding quality checking function. For better readability we provide semi-formal definitions only where necessary and refer to the online documentation and the implementation source code for further details. We also categorized the identified quality criteria into \emph{Labeling and Documentation Criteria}, \emph{Structural Criteria}, \emph{Linked Data Specific Criteria}, and \emph{Other Criteria}.

\todo{All}{decide for either using the term "concept" or "conceptual resource"}

% LEXICAL AND DOCUMENTATION CRITERIA

\subsection{Labeling and Documentation \textcolor{red}{Functions}}

The criteria defined in this section primarily focus on supporting interaction with human users that either work with the vocabulary itself or interact with a system using the vocabulary for, e.g., search and retrieval purposes.

\paragraph{QC 1: Omitted or Invalid Language Tags}

SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Since omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could limit language-dependent structural queries, we check a SKOS vocabulary for omitted language tags. We iterate over all resources in $IR$ and collect those that have relations to plain literals with missing or invalid language tags, hence tags that are not defined in RFC3066\footnote{RFC 3066: Tags for the Identification of Languages \url{http://www.ietf.org/rfc/rfc3066.txt}}.

% This is implementation specific; let's stay with high-level descriptions of the algorithms
%We delegate validity checking of language tags to methods provided by the by the OpenRDF\footnote{\url{http://www.openrdf.org/}} framework that in turn checks for conformance with RFC3066\footnote{\url{http://www.ietf.org/rfc/rfc3066.txt}}.\todo{CM}{is there any literature reference for this criterion?}

\paragraph{QC 2: Incomplete Language Coverage}
%A SKOS vocabulary that uses language tags with its literal values should apply them consistently for all concepts.
\cite{Soergel1997} points out the problem that some concepts may not be lexicalized in every language the vocabulary seeks to support. Presence of such concepts may indicate a need for restructuring the vocabulary or include documentation about the differing meanings in all supported languages. Hence, the set of language tags used by the literal values linked with a concept should be the same for all concepts. This is particularly important for applications that rely on internationalization and translation use cases. We calculate the affected concepts by extracting the global set of language tags used in a vocabulary from all literal values attached to all concepts in $C$. In a second pass we again iterate over this set of concepts and identify all conceptual resources having a set of language tags that is not equal to the global language set.

\paragraph{QC 3: Undocumented Concepts}

\cite{Svenonius1997} advocates the ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties''~\cite{SkosReference2008}. To identify all undocumented concepts, we iterate over all concepts in $C$ and collect those that do not use any of these documentation properties.

\paragraph{QC 4: Label Conflicts}

This criterion is an extension of a statement from \cite{Isaac2009} that recommends that ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Based on our review of existing vocabularies, we could observe that (i) many vocabularies don’t use concept schemes and (ii) identical alternative or hidden can also negatively affect use cases such as auto-completion, which proposes labels based on the user input. Thus, this criterion generalizes this statement by finding all concept pairs with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} property values meeting a certain similarity threshold defined by a function $sim:ULV \times ULV \rightarrow [0,1]$. The default, built-in similarity function checks for case-insensitive string equality with a threshold equal to 1. This criterion is calculated by iterating over all pairs of concepts, i.e., elements of $C \times C$, applying $sim$ to every possible label combination, and collecting those pairs for which $sim$ meets or exceeds a specified threshold.

>>>>

\section{Possible Quality Issues in SKOS Vocabularies}\label{sec:criteria}

@@

We identified an initial set of possible quality issues in SKOS vocabularies by reviewing existing literature and manually examining existing vocabularies. We published our findings online\footnote{\url{https://github.com/cmader/qSKOS/wiki/Quality-Criteria-for-SKOS-Vocabularies}} and requested feedback from experts via public mailing list and informal face to face discussions. Based on the received responses, we then selected a subset of these issues and formalized them into computable \textbf{quality checking functions}. Each function operates on a SKOS vocabulary, which we define, for the purpose of this work, as follows:

\begin{definition}[SKOS Vocabulary] Let a SKOS vocabulary be a tuple of the form $V = \langle IR, C, SR, ULV \rangle$, with \(IR = I_{CEXT}(\texttt{rdfs:Resource}^\mathcal{I})\) being the set of \textbf{resources}, \(C = I_{CEXT}(\texttt{skos:Concept}^\mathcal{I})\) being the set of \textbf{concepts}, \(SR = I_{EXT}(\texttt{skos:semanticRelation}^\mathcal{I})\) being the set of \textbf{semantic relations} associating conceptual resources with one another, and $LV \subseteq I_{CEXT}(\texttt{rdfs:Literal}^\mathcal{I})$ being the set of untyped \textbf{plain literals}. Further, we let $V$ be the fully entailed RDFS interpretation~\cite{RDFSEM2012} of the underlying RDF graph, enriched by entailment of \texttt{owl:inverseOf} properties.

@@

A quality checking function is applied on a given SKOS vocabulary and finds all vocabulary resources that match a certain issue and might degrade the overall quality of a vocabulary. Therefore, we can formalize a quality issue in terms of an abstract function $f:V \rightarrow \mathcal{P}(IR)$ that maps a given vocabulary $v \in V$ to the powerset of its resources. Each quality issue we found can be implemented as a realization of that abstract function and possibly take additional parameters. Statistical analysis can be computed on top of the result set.

In the following, we explain the origins and design rationale for each quality issue (QI) and define the corresponding quality checking function. For better readability and due to lack of space we provide only semi-formal definitions and refer to the online documentation and the implementation source code for further details. We also categorized the identified quality issues into \emph{Labeling and Documentation Issues}, \emph{Structural Issues}, \emph{Linked Data Specific Issues}, and \emph{Other Issues}.

% LABELING AND DOCUMENTATION ISSUES

\subsection{Labeling and Documentation Issues}

The following issues primarily focus on possible problems that might arise when human users work with a SKOS vocabulary or interact with a system that uses the vocabulary for, e.g., search and retrieval purposes.

\paragraph{QI 1: Omitted or Invalid Language Tags}

% Origin and rationale
SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could unintentionally limit the result set of language-dependent structural queries.
% Function defintion
We can check a SKOS vocabulary for omitted and invalid language tags by defining a quality checking function $oilt$ that iterates over all resources in $IR$ and finds those that have relations to plain literals with missing or invalid language tags, hence tags that are not defined in RFC3066\footnote{RFC 3066: Tags for the Identification of Languages \url{http://www.ietf.org/rfc/rfc3066.txt}}.

\paragraph{QI 2: Incomplete Language Coverage}

% Origin and rationale
Soergel et al.~\cite{Soergel1997} point out that ``the problem that has bedeviled many developers of multilingual thesauri is that a concept lexicalized in one language may not be lexicalized in another...''. In the context of SKOS, this means that the set of language tags used by the literal values linked with a concept should be the same for all concepts. This is particularly important for applications that rely on internationalization and translation use cases.
% Function defintion
We identify affected the concepts by a function $ilc$ that first extracts the global set of language tags used in a vocabulary from all literal values attached to all concepts in $C$. In a second iteration over all concepts it identifies all concepts having a set of language tags that is not equal to the global language set.

\paragraph{QI 3: Undocumented Concepts}

% Origin and rationale
Svenonius~\cite{Svenonius1997} advocates the ``inclusion of as much definition material as possible'' and the SKOS language specification~\cite{SkosReference2008} defines a set of properties, denoted as ``documentation properties'', that are intended to hold this kind of information.
% Function defintion
To identify all undocumented concepts, we define a function $uc$ we iterate over all concepts in $C$ and by default collect those that do not use any of these documentation properties.

\paragraph{QI 4: Label Conflicts}

% Origin and rationale
This issue is an extension of a statement from the SKOS Primer~\cite{Isaac2009} that recommends that ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. We reviewed existing vocabularies and applications and found that (i) many vocabularies don't use concept schemes and (ii) identical \texttt{skos:altLabels} or \texttt{skos:hiddenLabels} can also negatively affect use cases such as auto-completion, which proposes labels based on the user input.
% Function defintion
Thus, we generalize this statement and define a function $lc$ that finds all concept pairs with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} property values meeting a certain similarity threshold defined by a function $sim:ULV \times ULV \rightarrow [0,1]$. The default, built-in similarity function checks for case-insensitive string equality with a threshold equal to 1. The function iterates over all concept pairs $C \times C$, applies $sim$ to every possible label combination, and collects those pairs with at least one label combination that meet or exceed a specified similarity threshold.
